Six ways to understand monads\\ \small{Experiences from the functional programming course at TU Delft}

### 2012-11-01 latex/paper.tex
Teaching by example - done

The above led us to not choose any particular language for the course. 
During the lectures, the demonstration languages would be Haskell and
{\sc c\#}. We actively encouraged students to use the language of their
choice on the platform of their choice to carry out homework assignments
and the course's final project.

@@

The focus of teaching was to present functional programming from a 
practical aspect.

The proverbial ``six ways to understand monads'' included presentations of the
formal theoretical concepts, visualizations of monads as blenders (based on the
monadic property that once a value is captured, it is then contained),
applications in Haskell for encapsulating state (through the \textsf{Maybe}
monad), applications to generic programming in {\sc c\#}, applications to {\sc
linq} and the reactive extensions framework, and, finally, a whiteboard exercise
where students where prompted to participate to the specification of Haskell's
\textsf{IO} monad from basic principles.

\subsection{The teapot experiment}

@@

One of the highlights of the taught period of the course was the teapot
experiment. While not a formal experiment in the scientific sense of the word,
we used it to draw the student's attention to the following facts:

@@

The student's response was overwhelming. Even though it was made clear that the
exercise would not contribute to the final grade, the will to apply the data
processing techniques taught during the lectures motivated the students to work
very hard. The fact that they were instructed to use their favourite language
enabled them to focus on decomposing the problem in a series of testable
calculation steps rather than meddling with the intricacies of a pure functional
language.  Despite the advice, students were eager to test-drive the newly
acquired knowledge in a functional language; apart from {\sc c\#} and
Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
renderings produced by student programs can be seen in Figure~\ref{fig:teapots}.

\subsection{Teaching Map/Reduce}

According to many students, the Map/Reduce data analysis framework was what
drove them to attend the course. During the course, several examples
where presented

The result was that in their projects, students used implicit, through
the selected language's library collection methods, or explicit, by
formalizing data processing as a series of in memory Map/Reduce operations, forms
of Map/Reduce to process the data.

Out of personal interest, one student even implemented L\"ammel Map/Reduce
formulation in Scala, and through the language's extension mechanisms,
adapted it for use as an extension method by any Scala collection type,
including parallel collections.

@@

Coding sprints or ``hackathons'' have long been used by open source software
projects to speed up development and increase participation of community
members~\cite{Adams11}. In a typical sprint, project members set measurable
targets and work intensively in a tight deadline to deliver them. To ensure
the timely delivery of the student projects, we organized an one day
coding sprint, during the students where request to i) set an explicit target 
ii) write code for 7 hours with a short break iii) present a 5 minute 
demo of their work at the end of the day. All project teams where expected to
work in the same room, while communication among the teams was facilitated
by the lab demonstrator, to avoid unecessary distractions.

\subsection{Student Projects}

As expected, the performance of students in project varied.

@@

We would like to thank the course participants for their enthusiasm and
dedication. The source code to all student projects and the teapot
exercise can be found at (or linked from) \url{https://github.com/gousiosg/teapots}.

>>>>

The above led us to not choose any particular language for the course.  During
the lectures, the demonstration languages where Haskell and {\sc c\#}, while at
the labs we used Scala. We actively encouraged students to use the language of
their choice on the platform of their choice to carry out homework assignments
and the course's final project.

@@

The focus of teaching was to present functional programming from a practical
aspect; we were mostly interested to teach students what they can do rather than
the theory behind functional constructs. An important distinction
that was made early on was that the every day world is imperative; therefore
while functional programming can be a great tool for thinking about a problem, a
pure functional solution cannot reflect on the real world. As a consequence it
is usually best to mix programming paradigms, i.e. imperative or object-oriented
programming for state representation and functional programming for data
processing. In our experience, this distinction helped students understand
that they are already doing functional programming without noticing;
enforcing data immutability to handle multi-threading problems or passing
function arguments in scripting languages are applications of functional
programming principles. The moment students realized this fact, their
attention was captured to the lectures.

During both lectures and labs, all non-whiteboard examples where actually
entered in a real programming environment (LinqPad or the Scala command-line
read-evaluate loop) and the results of the evaluation where discussed with the
students. Participation was encouraged by modifying the examples and asking the
students to perform the evaluation, before the example was executed. The
examples varied from list manipulation with higher-order functions to 
a toy actor model example.

The proverbial ``six ways to understand monads'' included presentations of the
formal theoretical concepts, visualizations of monads as blenders (based on the
monadic property that once a value is captured, it is then contained),
applications in Haskell for encapsulating state (through the \textsf{Maybe}
monad), applications to generic programming in {\sc c\#}, applications to {\sc
linq} and the reactive extensions frameworks, and, finally, a whiteboard
exercise where students were prompted to participate in the ad-hoc specification
of Haskell's \textsf{IO} monad from basic principles. Once again, the examples
enabled the students to understand the composition and state containment
properties of an otherwise difficult to comprehend theoretical concept.

\subsection{The teapot exercise}

@@

One of the highlights of the taught period of the course was the teapot
exercise, which 
we used it to draw the student's attention to the following facts:

@@

The student's response was overwhelming. Even though it was made clear that the
exercise would not contribute to the final grade, the will to apply the data
processing techniques taught during the lectures motivated the students to work
very hard. The fact that they were instructed to use their favourite language
enabled them to focus on decomposing the problem in a series of testable
calculation steps rather than meddling with the intricacies of a pure functional
language. In spite of the advice, students were eager to test-drive the newly
acquired knowledge in a functional language; apart from {\sc c\#} and
Javascript, solutions were also provided in Scala, Scheme and Haskell. Example
renderings produced by student programs can be seen in Figure~\ref{fig:teapots}.

\subsection{Map/Reduce}

According to many students, the Map/Reduce data analysis framework was what
drove them to attend the course. The lecture on Map/Reduce featured an invited
talk by Ralf L\"ammel, who presented his formulation of Map/Reduce using
functional programming principles~\cite{Lamme08} and explained how those can be
applied to process semi-structured data. The talk resonated to students; in
their projects, students used implicit, through the selected language's library
collection methods, or explicit, by formalizing data processing as a series of
in memory Map/Reduce operations, forms of Map/Reduce to process their datasets.
Out of personal interest, one student even implemented L\"ammel Map/Reduce
formulation in Scala, and through the language's extension mechanisms, adapted
it for use as an extension method by any Scala collection type, including
parallel collections.

@@

Coding sprints or ``hackathons'' have long been used by open source software
projects to speed up development and increase participation of community
members~\cite{Adams11}. In a typical sprint, project members set measurable
targets and work intensively in a tight deadline to deliver them. To ensure
the timely delivery of the student projects, we organized an one day
coding sprint, during the students where request to i) set an explicit target 
ii) write code for 7 hours with a short break iii) present a 5 minute 
demo of their work at the end of the day. All project teams where expected to
work in the same room, while communication among the teams was facilitated
by the lab demonstrator, to avoid unnecessary distractions.

The sprint helped students kick start their projects.

\subsection{The student projects}

As expected, the performance of students in their projects varied.

@@

We would like to thank the course participants for their enthusiasm and
dedication. The source code to all student projects and the teapot
exercise can be found at (or linked from) \url{http://swerl.tudelft.nl/bin/view/Main/FunctionalProgrammingCourse}.
